---
title: "HW1-1"
author: "Zeqi Niu"
date: "10/27/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

## HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. To complete this assignment, follow these steps:

1. Download the `HW1.Rmd` file from Canvas. 

3. Open `HW1.Rmd` in RStudio. Replace the "Your Name Here" text in the `author:` field with your name.

4. Supply your solutions to the homework by editing `HW1.Rmd`.

5. Run your code in the Console and Knit HTML frequently to check for errors.

6. You may find it easier to solve a problem by interacting only with the Console at first. 

7. When you have completed the homework please check that your knits correctly when you click `Knit HTML`.

8. Submit your RMD file and the HTML output.


## Data frame basics

We will work with the nycflights dataset. **Please be sure to keep the data file in the same directory as the RMD file - your project directory.** and make that directory the working directory.

First - make sure that you are reading the data correctly and doing all the data cleaning steps that we did in class. Feel free to go beyond our class work in cleaning the data if you feel the need. 

```{r, cache = TRUE}
#Reading and exploring the data 
nyc <- read.csv("nycflights.csv")
head(nyc, n = 10)
str(nyc)
summary(nyc)
```

I now would like you to answer the following - all in their own separate R code blocks.

### Cleaning Data

```{r}
# Delte Unnessary columns
nyc$X <- NULL
nyc$year <- NULL
```

```{r}
# Factorize columns
#flight, carrier, origin, dest, month, tailnum
nyc$flight <- as.factor(nyc$flight)
nyc$carrier <- as.factor(nyc$carrier)
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)
nyc$month <- as.factor(nyc$month)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$flightdate <- as.Date(nyc$time_hour, format = "%Y-%m-%d %H:%M:%S")
```

### Data Exploration

Let's first do some simple exploration of this data. 

- How many airlines are there? (Hint: `levels` and `length` can be useful here)

```{r}
length(levels(nyc$carrier))
nlevels(nyc$carrier)
```

- How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)

```{r}
nrow(nyc[nyc$carrier == "OO",])
```

- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)

```{r}
min(nyc$air_time, na.rm = TRUE)
```

- How many flights where there by United Airlines (code: UA) on Jan 12th 2013?

```{r}
nrow(nyc[nyc$carrier == "UA" & nyc$month == 1 & nyc$day == 12,])
```

### Arrival Delay

Lets focus on Arrival Delay.

- What was the average arrival delay for all airports and all airlines combined in 2013? 

```{r}
round(mean(nyc$arr_delay, na.rm = TRUE))
```

- Whats was the median arrival delay for all airports and all airlines combined in 2013?

```{r}
median(nyc$arr_delay, na.rm = TRUE)
```

- Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

The distribution is positivly skewed, most flights are relatively on time, the data is skewed by minority of flights that was very delayed.  


### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

- Calculate average arrival delays by airline (Hint: look up the command `tapply`)

```{r}
arr_delay_by_airline <- tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)
round(arr_delay_by_airline, 2)
```


- Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)

```{r}
barplot(sort(arr_delay_by_airline), las = 2) # las the axis 
```


- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? Provide answer to this question in a text paragraph form using **inline R code**.

**`r names(arr_delay_by_airline[arr_delay_by_airline == max(arr_delay_by_airline)]) `** has the highest average arrival delay; **`r names(arr_delay_by_airline[arr_delay_by_airline == min(arr_delay_by_airline)]) `** has the smallest average arrival delay; airlines that actually have negative average delay are: **`r names(arr_delay_by_airline[arr_delay_by_airline < 0])`**

### Air Gain

Create a new column named airgain such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air.

```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
```


a) Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.

```{r}
summary(nyc$airgain)
hist(nyc$airgain)
barplot(sort(tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)), las = 2)
```

b) Answer the questions:

  - do airlines actually gain any time when in air on average? 
Yes, the average airgain is 5.66, meaning the depature delay is larger than arrival delay on average, airlines are gaining time in air. 
  - Calculate average airgain for different airlines - which airlines do a better job, which do a worse job? 
VX, HA, AS do a better job, F9, FL, MQ do a worse job. 


### Making Your HTML Look Nice

We want our report to be good looking, professional documents. To that end, I am asking you to do the following:

- Have a floating table of contents

- Include code folding in your output. You can find more about code folding here: https://bookdown.org/yihui/rmarkdown/html-document.html#code-folding


That's it. Once you are done, make sure everything works and knits well and then you can push your changes to the GitHub repo and uplaod the RMD flile and the html output to Canvas.

**Have Fun!**

Sanjeev
